---
title: "Analysis of Asthma Control"
author: "Tae Yoon (Harry) Lee"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(VGAM)
select <- dplyr::select
```

# Data source
Economic Burden of Asthma (EBA) data contains measurements of asthma control levels for 618 patients
(74% are >= 18 years old). The measurements were taken every 3 months for a year. We followed the
2020 GINA guidelines to define asthma control level by using the sum of the four indicator variables
(0 if no and 1 if yes): daily symptoms, nocturnal symptoms, inhaler use, and limited activities in
the last 3 months before each measurement. If the sum is zero, then the asthma control level is
controlled. If it is less than 3, then it is partially controlled. Otherwise, it is uncontrolled.
For responses with \textit{do not know} to the indicator variables, we treated them as a \textit{no}.
In this analysis, we did not consider treatment nor whether a patient experienced an exacerbation
in the last 3 months before the visit. We excluded two patients whose asthma diagnosis dates were
earlier than they were born and three patients who had no asthma diagnosis dates.

```{r balance, echo=FALSE, eval=FALSE}
df_demo <- haven::read_sas(here("private_dataset", "demo.sas7bdat")) %>%
    mutate(age=as.numeric(startDate - birthdate) / 365.25) %>% # age in years
    mutate(sex=if_else(gender=='f', 0, 1)) %>%  # female: 0; male: 1
    select(studyId, sex, age)
df_baseline <- haven::read_sas(here("private_dataset", "astat0.sas7bdat"))
df_fu <- haven::read_sas(here("private_dataset", "astatf.sas7bdat"))

df_EBA <- df_baseline %>%
    select(
        studyId, visit, daytimeSymptoms, nocturnalSymptoms, inhalerUse, limitedActivities,
        exacerbations
    )
df_EBA <- df_EBA %>%
    rbind(
        .,
        df_fu %>%
            select(
                studyId, visit, daytimeSymptoms, nocturnalSymptoms, inhalerUse, limitedActivities,
                exacerbations
            )
    ) %>%
df_EBA <- df_EBA %>%
    left_join(df_demo, by="studyId") %>%
    left_join(
        df_baseline %>% select(studyId, ageAtAsthmaDx),
        by="studyId"
    ) %>%
    mutate(
        control_score=(
            as.numeric(daytimeSymptoms==1) +
            as.numeric(nocturnalSymptoms==1) +
            as.numeric(inhalerUse==1) +
            as.numeric(limitedActivities==1)
        )
    ) %>%
    mutate(
        control_score_missing=(
            as.numeric(!(daytimeSymptoms %in% c(1, 2))) +
            as.numeric(!(nocturnalSymptoms %in% c(1, 2))) +
            as.numeric(!(inhalerUse %in% c(1, 2))) +
            as.numeric(!(limitedActivities %in% c(1, 2)))
        )
    )

# use the GINA guidelines to determine the control level (Box 2-2, GINA guidelines 2020)
df_EBA <- df_EBA %>%
    mutate(
        control_level=case_when(
            control_score == 0 ~ 1,
            control_score < 3 & control_score > 0 ~ 2,
            control_score >= 3 ~ 3
        )
    )
df_EBA <- df_EBA %>%
    mutate(
        studyid=studyId,
        exac=as.numeric(exacerbations==1)
    )

# asthma diagnosis before birth? => something wrong with the data
df_EBA <- df_EBA %>%
    filter(!(studyId %in% c("13205707", "23207811")))

# no asthma Dx date => ??? data quality issues omit them (3 patients)
# treat "don't know/missing" as "no" for the four asthma control indicators and exacerbation
df_EBA <- df_EBA %>%
    na.omit()

write_rds(df_EBA, here("R/private_dataset/df_EBA.rds"))
```

The `df_EBA` dataset contains the following variables:

| name        | type           | description |
| ------------- |:------| :---------|
| studyID   | str           | 8 numeric characters |
| visit      | int  |    |
| daytimeSymptoms |  int  | 1 = TODO, 2 = TODO |
| nocturnalSymptoms | int  | 1 = TODO, 2 = TODO |
| inhalerUse | int  | 1 = TODO, 2 = TODO |
| limitedActivities | int  | 1 = TODO, 2 = TODO |
| exacerbations | int  | 1 = TODO, 2 = TODO |
| sex | int  | 0 = female, 1 = male |
| age | float | age in years |
| ageAtAsthmaDx | float |  |
| control_score | int  | 0 = fully-controlled, 1 = partially-controlled, 2 = partially-controlled, 3 = uncontrolled, 4 = uncontrolled |
| control_score_missing | int  | 0 = TODO, 1 = TODO |
| control_level | int  | 1 = fully-controlled, 2 = partially-controlled, 3 = uncontrolled |
| time_since_Dx | float | years? |
| studyid | str | 8 numeric characters, copy of studyId |
| exac | int  | 0 = TODO, 1 = TODO |
| time_since_Dx_cat | int  | 1 = TODO, 2 = TODO, 3 = TODO |


```{r, echo = F}
df_EBA <- read_rds(here("R/private_dataset/df_EBA.rds"))
write_csv(
    df_EBA %>% select(-studyid, -exac, -control_score, -control_score_missing, -control_level),
    here("R/private_dataset/df_EBA.csv")
)
new_id <- 1:length(unique(df_EBA$studyId))
df_id <- data.frame(new_id=new_id, studyId=unique(df_EBA$studyId))

df_EBA_analysis <- df_EBA %>%
    left_join(df_id, by=c("studyId")) %>%
    select(-studyId) %>%
    rename(studyId=new_id) %>%
    mutate(studyid=studyId) %>%
    mutate(group=as.numeric(age >= 18)) %>% # 0 pediatric 1 adult
    mutate(
        sex=as.factor(sex),
        control_level=ifelse(control_level==1, 0, 1),
        control_level=as.ordered(control_level),
        studyid=as.factor(studyId),
        age=age / 100,
        age_sq=age^2
    ) %>%
    select(studyid, visit, group, control_level, sex, age, age_sq, exac)
```

# Descriptive analysis
Among 613 patients, only 6\% were lost during the one-year follow-up. There were at least 500 cases
for each asthma control level. More females were in the cohort as expected
(adult asthma is more prevalent among females). Asthma control level changed during the FU for
79\% of the patients. There were at least 500 cases for each control level.

# Modeling
We used a random-effects proportional ordinal logistic regression using the package
$\textit{ordinal}$. The model is:

$$
\begin{align}
\mbox{logit}(P(Y_i \leq j|X_i)) &= \theta_j + X_i^\top \beta - u_i, ~\mbox{for } j \in \{1,2\} \\
 \sum_{j}P(Y_i=j|X_i) &= 1
\end{align}
$$

where $\theta_j$ are the cut points or threshold parameters, $X_i$ is the covariates for
patient $i$, $\beta$ is the corresponding parameters, and $u_i \sim N(0,\sigma^2_u)$ are
patient-specific random-effects. The individual probability at level $j$ was computed as:

$$
P(Y_i=j) = P(Y_i\leq j) - P(Y_i \leq j-1).
$$

# Random-effects ordinal regression
We compared the two links, probit and logit, and chose the link that provided better goodness-of-fit
based on Akaike information criterion (the lower the better). The model with the logit link better
fitted the data (AIC = 5512) than that with the probit link (AIC=5522).

```{r, echo = F}
library(ordinal)
library(sure)
formula1 <- as.formula("control_level ~ sex*age + sex*age_sq")
model_logistic <- ordinal::clmm2(
    formula=formula1, random=studyid, data=df_EBA_analysis, Hess=TRUE, nAGQ=5, link="logistic"
)
stats::AIC(model_logistic)
# AIC: 5511.6
model_probit <- ordinal::clmm2(
    formula=formula1, random=studyid, data=df_EBA_analysis, Hess=TRUE, nAGQ=5, link="probit"
)
stats::AIC(model_probit)
# AIC: 5522.0;
# logistic better
summary(model_logistic)
```